---
title: "aiMeRA User's Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(aiMeRA)
```



## **Quick start**

-------------   

## **Examples of advanced use**

-------------  

## **Functions**

- **data.setup()**   
- **data2sdmean()**   
- **interval()**   
- **mra()**    
- **netgraph()**    
- **read.rules()**   




---

## **data.setup()**     
Data setup for Modular Response Analysis

### Description


 Checks for possible errors in tables of experimental data and in the perturbation matrix. Returns data
 properly formated for the mra function.


### Usage

```r
data.setup(obj, pert.tab = NULL, dec = ".")
```


### Arguments

Argument      |Description
------------- |----------------
```obj```     |     A character string specifying the path to a folder that contains the experimental data tables and the perturbation rules, or a list containing several data tables, or a single data table.
```pert.tab```     |     Optional. The perturbation rule table if not provided in ```obj```.
```dec```     |     The decimal separator as in ```base::read.csv``` (default is ".")

### Value


 If more than one data table are checked, then a list of matrices in the correct
 format for the mra function is returned. If a single table is checked, then a single matrix in the right format is returned.
 It also checks for correspondance with the perurbation rules table.


### Examples

```r 
 #Example using q-PCR data stored in the package files
 data=data.setup(list(estr1_A,estr1_B,estr2_A,estr2_B,estr3_A,estr3_B))
 #Examples using a gene network from Kholodenko et.al. PNAS 2002 figure a and b
 #Data was obtained from the given R matrix
 data.A=data.setup(gene.network.A)
 data.B=data.setup(gene.network.B)
 ``` 



---

## **data2sdmean()**      
Standard deviation and average calculation of a data set.

### Description


 This function computes the standard deviation and average values of a data set. Namely, if n replicates were generated,
 then n data tables should be provided to ```data.setup``` and ```data2sdmean``` computes the average and standard deviation
 of each cell over the n replicates.


### Usage

```r
data2sdmean(x)
```


### Arguments

Argument      |Description
------------- |----------------
```x```     |     A list containing 2 or more data tables.

### Value


 A list containing two matrices, the standard deviations and mean values for each variable in x.


### Examples

```r 
 data=data.setup(list(estr1_A,estr1_B,estr2_A,estr2_B,estr3_A,estr3_B))
 data2sdmean(data)
 ``` 


---

## **interval()**     
Confidence interval calculation for MRA connectivity coefficients

### Description


 A function that estimates confidence intervals of MRA connectivity coefficients based on replicates. It uses a bootstrap algorithm.


### Usage

```r
interval(tab,sd.tab,matp,n=10000,nrep=2)
```


### Arguments

Argument      |Description
------------- |----------------
```tab```     |     A data table containing the experimental data in the format output by ```data.setup```.
```sd.tab```     |     A data table containing the standard deviation values.
```matp```     |     A perturbation rule table, with rows corresponding to the MRA modules and columns to perturbations.
```n```     |     Number of samples for the bootstrap algorithm.
```nrep```     |     Number of replicates. Default is 2.

### Value


 A list containing the upper and lower values of the confidence interval of each connectivity coefficient.


### Examples

```r 
 data=data.setup(list(estr1_A,estr1_B,estr2_A,estr2_B,estr3_A,estr3_B))
 
 # we first average the technical replicates of each biological replicate
 # and then only keep the (averaged) biological replicates for the calculation
 CODE
 sd.mean=data2sdmean(data)
 rules=c("Et->Luciferase","E2+siRIP140->RIP140","E2+siLCoR->LCoR","E2->0")
 matp=read.rules(rules)
 interval(sd.mean$mean,sd.mean$sd,matp,nrep=6)
 ``` 



---

## **mra()**     
Modular Reponse Analysis function

### Description


 Calculation of connectivity coefficients between MRA modules.


### Usage

```r
mra(tab,matp,check=TRUE)
```


### Arguments

Argument      |Description
------------- |----------------
```tab```     |     Data in the format returned by ```data.setup``` or a single data table in the format required by ```data.setup```.
```matp```     |     A perturbation rule table, with rows corresponding to the MRA modules and columns to perturbations.
```check```     |     Logical. If ```TRUE``` then data and perturbation rule are checked for consistency (by calling ```data.setup```).

### Details


 It assumes that one perturbation must affect only one module of the network. This is specified as binary values in
 the perturbation matrix. It also assumes that row names in data tables are the names of biological modules and column names are the
 the names of perturbations.  ---->> A DEFINIR SEPAREMENT


### Value


 A list containing the connectivity map, the local responses matrix, the network responses matrix to perturbations
 and the basal line for all the modules.


### Examples

```r 
 #It creates the connectivity map between 2 transcriptional nuclear coregulators
 #(RIP140 and LCoR) and estrogen receptor alpha transcriptional activity reported by
 #a luciferase gene. q-PCR data is stored in the package files used below in the function data.setup
 #The model is obtained using the E2 stimulated condition.
 
 data=data.setup(list(estr1_A,estr1_B,estr2_A,estr2_B,estr3_A,estr3_B))
 sd.mean=data2sdmean(data)
 rules=c("Et->Luciferase","E2+siRIP140->RIP140","E2+siLCoR->LCoR","E2->0")
 matp=read.rules(rules)
 mra(sd.mean$mean,matp,check=TRUE)
 ``` 



---

## **netgraph()**    
Plot of network connectivity map

### Description


 Plots a graphic representation of a biological network connectivity map calculated by MRA.


### Usage

```r
netgraph(map,layout=igraph::layout_with_kk,pertu=FALSE,inter=NULL,
      cutoff=NULL,main=NULL,digits=2,v.type="circle",p.type="rectangle",
      outfile=NULL, ...)
```


### Arguments

Argument      |Description
------------- |----------------
```map```     |     A list containing the connectivity map (link matrix) of a network, and local matrix responses to perturbations, i.e., the output of ```mra```.
```layout```     |     Either a data.frame containing the x and y coordinates and the color of the vertices in the network, or a layout function according to the igraph package. Default uses the "layout_with_kk" igraph layout.
```pertu```     |     A boolean specifying whether perturbations should be plotted as vertices in the network.
```inter```     |     Confidence intervals calculated by ```interval```; connectivity coefficient with a confidence interval that does not include 0 are deemed significant and marked by an asterisk.
```cutoff```     |     Minimum value for a connectivity link coefficient between two modules for being plotted.
```main```     |     A character string giving the title of the plot.
```digits```     |     Number of digits to be plotted for each connectivity coefficient.
```v.type```     |     Vertex type for biological modules. One of “none”, “circle”, “square”, “csquare”, “rectangle” “crectangle”, “vrectangle”, “pie”, “raster”, or “sphere”.
```p.type```     |     Vertex type for perturbations. One of “none”, “circle”, “square”, “csquare”, “rectangle” “crectangle”, “vrectangle”, “pie”, “raster”, or “sphere”.
```outfile```     |     Optional. A character string giving the name of the output file in graphML format.
```...```     |     Arguments to be passed to igraph plot such as vertex and edges plotting parameters.

### Value


 If layout is an igraph function, then a data.frame containing the coordinates of the layout selected is returned.
 If outfile is not NULL, then the network is written in the graphML file format.


### Examples

```r 
 data=data.setup(list(estr1_A,estr1_B,estr2_A,estr2_B,estr3_A,estr3_B))
 sd.mean=data2sdmean(data)
 rules=c("Et->Luciferase","E2+siRIP140->RIP140","E2+siLCoR->LCoR","E2->0")
 matp=read.rules(rules)
 map=mra(sd.mean$mean,matp,check=FALSE)
 inter=interval(sd.mean$mean,sd.mean$sd,matp,nrep=6)
 netgraph(map,pertu=TRUE,inter=inter)
 ``` 



---

## **read.rules()**   
Creates a perturbation matrix from a set of perturbation rules.

### Description


 A function that takes a set of perturbation rules as input and creates a perturbation matrix to be used in other functions of the package.


### Usage

```r
read.rules(obj)
```


### Arguments

Argument      |Description
------------- |----------------
```obj```     |     Either the path to a perturbation rules file without header, or a vector of strings specifying the perturbation rules (see details).

### Details


 Perturbation rules are a set of strings specifying the action of perturbations upon modules in a network. In both cases, modules
 and perturbations must have the same names as the row and column names in experimental data tables (names are case sensitive).
 
 The rules syntax is  ```Perturbation->Module``` for a single perturbation,```Perturbation->0``` to specify the basal line and
 ```0->Module``` to specify modules that were not perturbated, but for which connectivity issued from other modules exist (unidirectional MRA).
 At least two rules and a basal line must be defined.


### Examples

```r 
 rules=c("Et->Luciferase","E2+siRIP140->RIP140","E2+siLCoR->LCoR","E2->0")
 ``` 






